---
title: "Predominance groups analysis and Bifidobacteria prevalences and abundances"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(stringsAsFactors = FALSE)
```

#Prepare environment.
```{r}
library(tidyverse)
library(ggplot2)
library("reshape2")
library("grid")
library(data.table)
library(stringr)
library(ggalluvial)
library(ggpattern)
library(ggpubr)
library(rstatix)
```

## Load metadata table
```{r}
#metaphlan output table
mpa_file <- read_delim("data/species_composition/milk_infant_samples_merged_abundance_contam_rm.txt", delim = '\t')
#metadata table
metadata <- read_delim("data/metadata_complete.csv", delim = ',')
metadata <- metadata %>% select(-Correct_sample_name) 
```

# Prepare files for Bifidobacteria beta diversity plots and predominance groups assignment
## Get bifidobacteria mean relative abundance for each sample for beta diversity plot
```{r}
#separate full clade name into taxonomic levels
mpa_file.delim <- mpa_file %>% 
  tidyr::separate(clade_name, c('Kingdom', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species', 'Strain'), sep ='\\|', remove = FALSE, fill="right") 

#select clades assigned to species level
mpa_file.target <- mpa_file.delim %>% 
  filter(!is.na(Species) & is.na(Strain)) %>% 
  filter(stringr::str_detect(Species, 'coli') |
           stringr::str_detect(Species, 'fragilis') |
           stringr::str_detect(Species, 'pneumoniae') |
           stringr::str_detect(Species, 'michiganensis') |
           stringr::str_detect(Species, 'vulgatus') |
           stringr::str_detect(Species, 'parvula') |
           stringr::str_detect(Species, 'perfringens')|
           stringr::str_detect(Species, 'atypica')|
           stringr::str_detect(Species, 'dispar'))

#keep simplified species name only
drop <- c("clade_name", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus","Strain")
mpa_file.target = mpa_file.target[,!(names(mpa_file.target) %in% drop)]
mpa_file.target$Species <- str_remove(mpa_file.target$Species, pattern = "s__")
mpa_file.target$Species <- gsub("_", " ", mpa_file.target$Species)

mpa_file.target.long <- mpa_file.target %>% 
  pivot_longer(cols = !Species, names_to = "Sample", values_to = "Abundance") 

#split sample ID for later merge with metadata table
mpa_file.target.long <- mpa_file.target.long %>% 
  mutate(full_ID=Sample) %>% 
  tidyr::separate(Sample, c('Sample', 'number'), sep ='_', remove = FALSE, fill="right") 

#merge with metadata table
merged.target <- left_join(mpa_file.target.long, metadata, by=c("full_ID"="Full_name"))

merged.target.clean <- merged.target %>% 
  select(Species, Sample, Abundance, Sample_type) %>% 
  filter(!is.na(Sample_type))

#identify species of interest for biplot beta diversity
#mean calculated when bifido species are present (zeros are NAs)
#replace zeros with NA
merged.target.clean.na <- merged.target.clean
merged.target.clean.na[merged.target.clean.na == 0.00000] <- NA

merged.target.clean.meanRelAb <- merged.target.clean.na %>% 
  group_by(Sample) %>% 
  summarise(mean_rel_ab = mean(Abundance, na.rm=TRUE))

#replace NA with 0s
is.nan.data.frame <- function(x)
do.call(cbind, lapply(x, is.nan))
merged.target.clean.meanRelAb[is.nan(merged.target.clean.meanRelAb)] <- 0.00000
write_csv(merged.target.clean.meanRelAb, "data/predominance_group/bifidos_relabs_mean.csv")
```

## Define predominance groups for each sample
```{r}
mpa_file.delim.clean <- mpa_file.delim %>% 
  filter(!is.na(Species) & is.na(Strain)) %>% 
  select(-clade_name, -Kingdom, -Phylum, -Class, -Order, -Family, -Genus, -Strain)
  
mpa_file.delim.clean$Species <- str_remove(mpa_file.delim.clean$Species, pattern = "s__")

mpa_file.delim.clean.long <- mpa_file.delim.clean %>% 
  pivot_longer(cols=-Species,
               names_to = "Sample",
               values_to = "Abundance")
 
mpa_file.delim.clean.long.groups <- mpa_file.delim.clean.long %>% 
  group_by(Sample) %>% 
  arrange(desc(Abundance)) %>% 
  slice(1:1) %>% 
  mutate(Group=ifelse(Species=="Bifidobacterium_longum", "Group_1", 
                      ifelse(Species=="Bifidobacterium_breve", "Group_2",
                             ifelse(Species=="Bifidobacterium_bifidum", "Group_3", "Group_4")))) %>% 
  select(Group, Sample)

write_csv(mpa_file.delim.clean.long.groups, "data/predominance_group/bifidos_groups.csv")
```

# Prevalence predominance groups

## Manipulate species output
```{r}
#select clades assigned to species level
mpa_file.bifidos <- mpa_file.delim %>% 
  filter(!is.na(Species) & is.na(Strain)) %>% 
  filter(stringr::str_detect(Species, 'Bifido'))

#keep simplified species name only
drop <- c("clade_name", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus","Strain")
mpa_file.bifidos = mpa_file.bifidos[,!(names(mpa_file.bifidos) %in% drop)]
mpa_file.bifidos$Species <- str_remove(mpa_file.bifidos$Species, pattern = "s__")
mpa_file.bifidos$Species <- gsub("_", " ", mpa_file.bifidos$Species)

mpa_file.bifidos.long <- mpa_file.bifidos %>% 
  pivot_longer(cols = !Species, names_to = "Sample", values_to = "Abundance") 

#split sample ID for later merge with metadata table
mpa_file.bifidos.long <- mpa_file.bifidos.long %>% 
  mutate(full_ID=Sample) %>% 
  tidyr::separate(Sample, c('Sample', 'number'), sep ='_', remove = FALSE, fill="right") 
```

## Prevalence of each predominance group across milk and stool samples
```{r}
metadata.clean <- metadata %>% 
  select(Full_name, Bodysite) %>% 
  filter(!is.na(Bodysite))

mpa_file.delim.clean.long.metadata <- left_join(mpa_file.delim.clean.long, metadata.clean, by=c("Sample"="Full_name"))
  
mpa_file.delim.clean.long.metadata.stools <- mpa_file.delim.clean.long.metadata %>% 
  filter(Bodysite=="Stool") %>% 
  group_by(Sample, Bodysite) %>% 
  arrange(desc(Abundance)) %>% 
  slice(1:1) %>% 
  mutate(Group=ifelse(Species=="Bifidobacterium_longum", "Group_1", 
                      ifelse(Species=="Bifidobacterium_breve", "Group_2",
                             ifelse(Species=="Bifidobacterium_bifidum", "Group_3", "Group_4")))) %>% 
  select(Group, Sample, Abundance) %>% 
  group_by(Group,Bodysite) %>% 
  mutate(perc=n()*100/334) %>% 
  select(Group, perc, Bodysite) %>% 
  distinct() %>% 
  mutate(x=0)

mpa_file.delim.clean.long.metadata.milk <- mpa_file.delim.clean.long.metadata %>% 
  filter(Bodysite=="Milk") %>% 
  group_by(Sample, Bodysite) %>% 
  arrange(desc(Abundance)) %>% 
  slice(1:1) %>% 
  mutate(Group=ifelse(Species=="Bifidobacterium_longum", "Group_1", 
                      ifelse(Species=="Bifidobacterium_breve", "Group_2",
                             ifelse(Species=="Bifidobacterium_bifidum", "Group_3", "Group_4")))) %>% 
  select(Group, Sample, Abundance) %>% 
  group_by(Group,Bodysite) %>% 
  mutate(perc=n()*100/173) %>% 
  select(Group, perc, Bodysite) %>% 
  distinct() %>% 
  mutate(x=0)

mpa_file.delim.clean.long.metadata.milk.stool <- rbind(mpa_file.delim.clean.long.metadata.milk, mpa_file.delim.clean.long.metadata.stools)

bars.bifidos <- ggplot(mpa_file.delim.clean.long.metadata.milk.stool, aes(fill=Group, y=perc, x=x)) + 
  geom_bar(position="fill", stat="identity") +
  scale_fill_manual(values = c("#630a4c","#ad3691", "#de9ecd", "#f2e6ef"))+
  theme_bw() +
  facet_grid(~Bodysite)

pdf(file=paste0("figures/groups_stability/groups_stackedBar.pdf"), width=3.5, height=3)
bars.bifidos
invisible(dev.off())

bars.bifidos
```

# Alluvion plot predominance group
```{r}
mpa_file.delim.clean.long.groups.metadata <- left_join(mpa_file.delim.clean.long.groups, metadata, by=c("Sample"="Full_name"))

mpa_file.delim.clean.long.groups.metadata <- mpa_file.delim.clean.long.groups.metadata %>% 
  select(Group, Sample,Bodysite, Timepoint, `Mother Infant Pair`)

mpa_file.delim.clean.long.groups.metadata.milk <- mpa_file.delim.clean.long.groups.metadata %>% 
  filter(Bodysite=="Milk") 

couples_both_timepoints.milk <- c(mpa_file.delim.clean.long.groups.metadata.milk %>% 
  group_by(`Mother Infant Pair`) %>% 
  summarise(n=n()) %>% 
  filter(n>1) %>% 
  select(`Mother Infant Pair`) %>% 
  pull())

mpa_file.delim.clean.long.groups.metadata.milk.bothT <- mpa_file.delim.clean.long.groups.metadata.milk %>% 
  filter(`Mother Infant Pair` %in% couples_both_timepoints.milk)

mpa_file.delim.clean.long.groups.metadata.stool <- mpa_file.delim.clean.long.groups.metadata %>% 
  filter(Bodysite=="Stool") 

couples_both_timepoints.stool <- c(mpa_file.delim.clean.long.groups.metadata.stool %>% 
  group_by(`Mother Infant Pair`) %>% 
  summarise(n=n()) %>% 
  filter(n>1) %>% 
  select(`Mother Infant Pair`) %>% 
  pull())

mpa_file.delim.clean.long.groups.metadata.stool.bothT <- mpa_file.delim.clean.long.groups.metadata.stool %>% 
  filter(`Mother Infant Pair` %in% couples_both_timepoints.stool)
```

## Alluvion Plot for Milk
```{r}
al_milk <- read_delim("alluvion_milk.csv", delim = ',')

al_milk.clean <- al_milk %>% 
  mutate(tmp="tmp")

al_milk.clean.plot <- ggplot(data = al_milk.clean,
       aes(axis1 = T1, axis2 = T2, y = Perc)) +
  scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.2, .05)) +
  xlab("Demographic") +
  geom_alluvium(aes(fill = T1), alpha=1) +
  scale_x_discrete(expand = c(.1, .1)) +
  geom_flow() +
  geom_stratum(alpha = .3) +
  theme_bw()+
  scale_fill_manual(values = c("#630a4c","#ad3691", "#de9ecd", "#f2e6ef")) 

pdf(file=paste0("figures/alluvion_milk.pdf"), width=3.2, height=2.5)
al_milk.clean.plot
invisible(dev.off())

al_milk.clean.plot
```

## Alluvion Plot for Stool samples
```{r}
#STOOL
al_stool <- read_delim("alluvion_stool.csv", delim = ',')

al_stool.clean <- al_stool %>% 
  mutate(tmp="tmp")

al_stool.clean.plot <- ggplot(data = al_stool.clean,
       aes(axis1 = T1, axis2 = T2, y = Perc)) +
  scale_x_discrete(limits = c("Class", "Sex", "Age"), expand = c(.2, .05)) +
  xlab("Demographic") +
  geom_alluvium(aes(fill = T1), alpha=1) +
  scale_x_discrete(expand = c(.1, .1)) +
  geom_flow() +
  geom_stratum(alpha = .3) +
  theme_bw()+
  scale_fill_manual(values = c("#630a4c","#ad3691", "#de9ecd", "#f2e6ef")) 

pdf(file=paste0("figures/groups_stability/alluvion_stool.pdf"), width=3.2, height=2.5)
al_stool.clean.plot
invisible(dev.off())

al_stool.clean.plot
```

# Prevalences/Abundances Bifidobacteria
```{r}
#select clades assigned to species level
mpa_file.bifidos <- mpa_file.delim %>% 
  filter(!is.na(Species) & is.na(Strain)) %>% 
  filter(stringr::str_detect(Species, 'Bifido') | stringr::str_detect(Species, 'coli') 
         | stringr::str_detect(Species, 'fragilis') | stringr::str_detect(Species, 'Veillonella')
         | stringr::str_detect(Species, 'perfringens') | stringr::str_detect(Species, 'Phocaeiola')
         | stringr::str_detect(Species, 'coli') | stringr::str_detect(Species, 'Klebsiella'))

#keep simplified species name only
drop <- c("clade_name", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus","Strain")
mpa_file.bifidos = mpa_file.bifidos[,!(names(mpa_file.bifidos) %in% drop)]
mpa_file.bifidos$Species <- str_remove(mpa_file.bifidos$Species, pattern = "s__")
mpa_file.bifidos$Species <- gsub("_", " ", mpa_file.bifidos$Species)

mpa_file.bifidos.long <- mpa_file.bifidos %>% 
  pivot_longer(cols = !Species, names_to = "Sample", values_to = "Abundance") 

#split sample ID for later merge with metadata table
mpa_file.bifidos.long <- mpa_file.bifidos.long %>% 
  mutate(full_ID=Sample) %>% 
  tidyr::separate(Sample, c('Sample', 'number'), sep ='_', remove = FALSE, fill="right") 

merged.bifidos <- left_join(mpa_file.bifidos.long, metadata, by=c("full_ID"="Full_name"))
```

## Prevalence of bifidobacteria in breastfeeding at 6M vs not breastfeeding
```{r}

#select only B.longum, B.breve and B.bifidum in infant stool samples
bf <- merged.bifidos %>% 
  filter(!is.na(Exclusive_bf_6m)) %>% 
  filter(Species=="Bifidobacterium longum" | Species=="Bifidobacterium bifidum" | 
           Species=="Bifidobacterium breve") %>% 
  filter(Sample_type=="Stool_1M" | Sample_type=="Stool_6M") %>% 
  mutate(bf_6M=ifelse(Sample_type=="Stool_1M","1M",Exclusive_bf_6m)) %>% 
  group_by(Sample, bf_6M) %>% 
  mutate(sum=sum(Abundance))

#1 month samples
bf.clean.prevs.1M <- bf %>%
  filter(Sample_type=="Stool_1M") %>% 
  select(Species, Sample, Abundance) %>% 
  filter(Abundance>0) %>% 
  select(-Abundance) %>% 
  group_by(Species) %>% 
  summarize(n=n()) %>% 
  mutate(perc=n*100/175) %>% 
  mutate(Exclusive_bf_6m="1M", Sample_type="Stool_1M")

#6 months samples
bf.clean.prevs.6M <- bf %>%
  filter(Sample_type=="Stool_6M") %>% 
  select(Species, Sample, Abundance, Sample_type, Exclusive_bf_6m) %>% 
  filter(Abundance>0 & !is.na(Exclusive_bf_6m)) %>% 
  select(-Abundance) %>% 
  group_by(Species, Exclusive_bf_6m) %>% 
  summarize(n=n()) %>% 
  mutate(perc=n*100/159) %>% 
  mutate(Sample_type="Stool_6M")

bf.clean.prevs.allT <- rbind(bf.clean.prevs.1M, bf.clean.prevs.6M)

#reorder bars within groups
bf.clean.prevs.allT$Exclusive_bf_6m <- factor(bf.clean.prevs.allT$Exclusive_bf_6m, levels = c('1M', 'Yes', 'No'))
  
#plot prevalences
bf.prev.plot <- ggplot(data=bf.clean.prevs.allT, aes(x=reorder(Species, -perc), y=perc, fill=Exclusive_bf_6m)) +
geom_bar(stat="identity", position="dodge", colour="black") +
theme_bw()+
scale_fill_manual(values = c("#5e3c99", "#b2abd2","#c3abc9"))

pdf(file=paste0("figures/groups_stability/bifido_prevalence_bf6M.pdf"), width=4.5, height=2)
bf.prev.plot
invisible(dev.off())

bf.prev.plot
```

## Abundance of bifidobacteria in breastfeeding at 6M vs not breastfeeding
```{r}
bf.clean <- bf %>% 
  select(Species, Sample, Abundance, Sample_type, bf_6M) %>% 
  filter(!is.na(Sample_type) & Abundance>0)

anno_df = compare_means(Abundance ~ bf_6M, group.by = "Species", data = bf.clean) %>%
 mutate(y_pos = 40)

#reorder bars within groups
bf.clean$bf_6M <- factor(bf.clean$bf_6M, levels = c('1M', 'No', 'Yes'))

stat.test <- bf.clean %>%
  group_by(Species) %>%
  wilcox_test(Abundance ~ bf_6M) %>%
  adjust_pvalue(method = "bonferroni") %>%
  add_significance("p.adj")
stat.test

bf.abs.plot <- ggplot(bf.clean, aes(x=reorder(Species, -Abundance), y=Abundance, fill=bf_6M)) + 
  geom_boxplot(position=position_dodge()) +
  theme_light() +
  scale_fill_manual(values = c("#5e3c99", "#c3abc9","#b2abd2"))+
  stat_summary(fun.y = mean, color = "black", position = position_dodge(0.75),
             geom = "point", shape = 18, size = 3,
             show.legend = FALSE)

  stat_compare_means(aes(group = bf_6M), label = "p.signif")

pdf(file=paste0("figures/groups_stability/bifido_abundance_bf6M.pdf"), width=4, height=4)
bf.abs.plot
invisible(dev.off())

bf.abs.plot
```
