---
title: "Strain sharing among unrelated infants"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)
options(stringsAsFactors = FALSE)
```

#Prepare environment.
```{r}
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(gridExtra)
```

#Load tables
```{r}
#strainphlan table
strain_file <- read_delim("data/strains/strain_x_sample.csv", delim = ',')
#metadata
metadata <- read_delim("data/metadata.csv", delim = ',')
```

## Iterate over every infant - 1 months (T1)
```{r}
strain_file.samples <- colnames(strain_file)

stool.1M <- c(metadata %>% 
  filter(Sample_type=="Stool_1M") %>% 
  select(Old_sample_name) %>% 
  pull())

strain_file.stool1M <- intersect(strain_file.samples, stool.1M)
infants_loop.1M <- data.frame()

marked_asDone.1M <- list()

for (sample.name1.1M in strain_file.stool1M) {
  for (sample.name2.1M in strain_file.stool1M) {
      
    if (sample.name1.1M != sample.name2.1M) {
      
      s2s1.1M <- paste0(sample.name2.1M, sample.name1.1M)
      #create list of inverted sample pairs 
      marked_asDone.1M <- c(marked_asDone.1M, s2s1.1M)
      s1s2.1M <- paste0(sample.name1.1M, sample.name2.1M)

      #avoid calculation if s1s2 was already encountered previously as s2s1
        if (!(s1s2.1M %in% marked_asDone.1M)){
          
          strain_file.small.1M <- strain_file %>%
            select(sample.name1.1M, sample.name2.1M)
          
          n_strains_shared.1M <- length(strain_file.small.1M %>% 
            mutate(sum = rowSums(.)) %>% 
            filter(sum>1) %>% #present in both samples
            pull(sum))

          infants_loop.1M <- rbind(
            infants_loop.1M,
            data.frame(
              sample1 = sample.name1.1M,
              sample2 = sample.name2.1M,
              n_strains_shared_btw_infant_pairs = n_strains_shared.1M))
        }
      }

  }
}

#save to avoid recomputing the table (estimated time 7 min)
#write.csv(infants_loop.1M, "infants_loop_1M.csv") 
```

## Iterate over every infant - 6 months (T2)
```{r}
stool.6M <- c(metadata %>% 
  filter(Sample_type=="Stool_6M") %>% 
  select(Old_sample_name) %>% 
  pull())

strain_file.stool6M <- intersect(strain_file.samples, stool.6M)
infants_loop.6M <- data.frame()

marked_asDone.6M <- list()

for (sample.name1.6M in strain_file.stool6M) {
  for (sample.name2.6M in strain_file.stool6M) {
      
    if (sample.name1.6M != sample.name2.6M) {
      
      s2s1.6M <- paste0(sample.name2.6M, sample.name1.6M)
      #create list of inverted sample pairs 
      marked_asDone.6M <- c(marked_asDone.6M, s2s1.6M)
      s1s2.6M <- paste0(sample.name1.6M, sample.name2.6M)

      #avoid calculation if s1s2 was already encountered previously as s2s1
        if (!(s1s2.6M %in% marked_asDone.6M)){
          
          strain_file.small.6M <- strain_file %>%
            select(sample.name1.6M, sample.name2.6M)
          
          n_strains_shared.6M <- length(strain_file.small.6M %>% 
            mutate(sum = rowSums(.)) %>% 
            filter(sum>1) %>% #present in both samples
            pull(sum))

          infants_loop.6M <- rbind(
            infants_loop.6M,
            data.frame(
              sample1 = sample.name1.6M,
              sample2 = sample.name2.6M,
              n_strains_shared_btw_infant_pairs = n_strains_shared.6M))
        }
      }

  }
}

#save to avoid recomputing the table (estimated time 7 min)
#write.csv(infants_loop.6M, "infants_loop_6M.csv") 
```

# Stool samples 1 month

## add metadata 
```{r}
infants_loop.T1 <- read_delim("infants_loop_1M.csv", delim = ',')

#does each pair of unrelated infants at 1M share AT LEAST ONE strain?
infants_loop.T1.binary <- infants_loop.T1 %>% 
  mutate(atLeast_1strain_shared = ifelse(n_strains_shared_btw_infant_pairs>0, 1, 0)) %>% 
  select(-n_strains_shared_btw_infant_pairs, -`...1`)
  
metadata.clean1 <- metadata %>% 
  select(Old_sample_name, `Mother Infant Pair`, Place_birth, `Month/Year`) %>% 
  mutate(Old_sample_name1=Old_sample_name,
         `Mother Infant Pair1`=`Mother Infant Pair`,
         Place_birth1=Place_birth,
         `Month/Year1`=`Month/Year`) %>% 
  select(-Old_sample_name, -`Mother Infant Pair`, -Place_birth, -`Month/Year`) 

metadata.clean2 <- metadata %>% 
  select(Old_sample_name, `Mother Infant Pair`, Place_birth, `Month/Year`) %>% 
  mutate(Old_sample_name2=Old_sample_name,
         `Mother Infant Pair2`=`Mother Infant Pair`,
         Place_birth2=Place_birth,
         `Month/Year2`=`Month/Year`) %>% 
  select(-Old_sample_name, -`Mother Infant Pair`, -Place_birth, -`Month/Year`) 
  
infants_loop.T1.binary.hospYear <- left_join(infants_loop.T1.binary, metadata.clean1, by=c("sample1"="Old_sample_name1"))
infants_loop.T1.binary.hospYear.extended <- left_join(infants_loop.T1.binary.hospYear, metadata.clean2, by=c("sample2"="Old_sample_name2"))

infants_loop.T1.binary.hospYear.extended <- infants_loop.T1.binary.hospYear.extended %>% 
  separate(`Month/Year1`, c("Month1", "Year1"), sep = "_", remove=FALSE) %>% 
  separate(`Month/Year2`, c("Month2", "Year2"), sep = "_", remove=FALSE)

#all unrelated infants
infants_loop.T1.binary.hospYear.extended.clean <- infants_loop.T1.binary.hospYear.extended %>% 
  filter(!is.na(Place_birth1) & !is.na(Place_birth2)) %>% 
  mutate(location=ifelse(Place_birth1==Place_birth2, "Same_Hospital", "Diff_Hospital")) %>% 
  mutate(location_time= case_when(
    (location=="Same_Hospital" & `Month/Year1`==`Month/Year2`) ~ "SameHospital-SameMY",
    (location=="Same_Hospital" & `Month/Year1`!=`Month/Year2`) ~ "SameHospital-DiffMY",
    (location=="Diff_Hospital" & `Month/Year1`==`Month/Year2`) ~ "DiffHospital-SameMY",
    (location=="Diff_Hospital" & `Month/Year1`!=`Month/Year2`) ~ "DiffHospital-DiffMY",
    TRUE ~ "Other")) %>% 
    mutate(location_timeYearOnly= case_when(
    (location=="Same_Hospital" & `Year1`==`Year2`) ~ "SameHospital-SameY",
    (location=="Same_Hospital" & `Year1`!=`Year2`) ~ "SameHospital-DiffY",
    (location=="Diff_Hospital" & `Year1`==`Year2`) ~ "DiffHospital-SameY",
    (location=="Diff_Hospital" & `Year1`!=`Year2`) ~ "DiffHospital-DiffY",
    TRUE ~ "Other"))

infants_loop.T1.binary.hospYear.extended.clean$atLeast_1strain_shared <- as.integer(infants_loop.T1.binary.hospYear.extended.clean$atLeast_1strain_shared)
```

# Stool samples 1 month (T1) - Fisher exact tests

## Hospital Only
```{r}
fisher.df.location.T1 <- infants_loop.T1.binary.hospYear.extended.clean %>% 
  group_by(atLeast_1strain_shared, location) %>% 
  summarise(counts=n()) %>% 
  mutate(atLeast_1strain_shared2 = ifelse(atLeast_1strain_shared==0, "No", "Yes")) %>% 
  ungroup() %>% 
  select(-atLeast_1strain_shared)

#df for fisher exact test
fisher.df.location.T1.wide <- as.data.frame(fisher.df.location.T1 %>% 
  pivot_wider(names_from = atLeast_1strain_shared2,
              values_from = counts) %>% 
  select(location,Yes, No) %>% 
  slice(2,1))

rownames(fisher.df.location.T1.wide) <- fisher.df.location.T1.wide[,1]
fisher.df.location.T1.wide[,1] <- NULL

fisher.test(fisher.df.location.T1.wide)
#Hospital Only
```

## Hospital + Year
```{r}
fisher.df.locationY.T1 <- infants_loop.T1.binary.hospYear.extended.clean %>% 
  group_by(atLeast_1strain_shared, location_timeYearOnly) %>% 
  summarise(counts=n()) %>% 
  mutate(atLeast_1strain_shared2 = ifelse(atLeast_1strain_shared==0, "No", "Yes")) %>% 
  ungroup() %>% 
  select(-atLeast_1strain_shared)

#df for fisher exact test
fisher.df.location.T1.wideY <- as.data.frame(fisher.df.locationY.T1 %>% 
  pivot_wider(names_from = atLeast_1strain_shared2,
              values_from = counts) %>% 
  select(location_timeYearOnly,Yes, No) %>% 
  slice(2,1))

rownames(fisher.df.location.T1.wideY) <- fisher.df.location.T1.wideY[,1]
fisher.df.location.T1.wideY[,1] <- NULL

fisher.test(fisher.df.location.T1.wideY)
#hospital and Year
```

## Barplot from fisher df
```{r}
#SAME HOSPITAL---------
signal.sameH_sharingYes_num.T1 <- fisher.df.location.T1.wide[1,1]
signal.sameH_sharingAll_den.T1 <- fisher.df.location.T1.wide[1,2]+fisher.df.location.T1.wide[1,1]

#baseline
baseline.sameH_sharingYes_num.T1 <- fisher.df.location.T1.wide[2,1]
baseline.sameH_sharingAll_den.T1 <- fisher.df.location.T1.wide[2,2] + fisher.df.location.T1.wide[2,1]

signal.sameH.T1 <- signal.sameH_sharingYes_num.T1/signal.sameH_sharingAll_den.T1
baseline.sameH.T1 <- baseline.sameH_sharingYes_num.T1/baseline.sameH_sharingAll_den.T1

#SAME HOSPITAL + YEAR---------
signal.sameHY_sharingYes_num.T1 <- fisher.df.location.T1.wideY[1,1]
signal.sameHY_sharingAll_den.T1 <- fisher.df.location.T1.wideY[1,2] + fisher.df.location.T1.wideY[1,1]
baseline.sameHY_sharingYes_num.T1 <- fisher.df.location.T1.wideY[2,1]
baseline.sameHY_sharingAll_den.T1 <- fisher.df.location.T1.wideY[2,2] + fisher.df.location.T1.wideY[2,1]

signal.sameHY.T1 <- signal.sameHY_sharingYes_num.T1/signal.sameHY_sharingAll_den.T1
baseline.sameHY.T1 <- baseline.sameHY_sharingYes_num.T1/baseline.sameHY_sharingAll_den.T1

Timepoint.T1 <- c("T1", "T1")
Condition.T1 <- c("SameH", "SameHY")
signal.T1 <- c(signal.sameH.T1, signal.sameHY.T1)
baseline.T1 <- c(baseline.sameH.T1, baseline.sameHY.T1)

values_df.T1 <- data.frame(Timepoint.T1, Condition.T1, signal.T1, baseline.T1)
```

# Stool samples 6 months

## Add metadata  
```{r}
infants_loop.T2 <- read_delim("infants_loop_6M.csv", delim = ',')

#does each pair of unrelated infants at 1M share AT LEAST ONE strain?
infants_loop.T2.binary <- infants_loop.T2 %>% 
  mutate(atLeast_1strain_shared = ifelse(n_strains_shared_btw_infant_pairs>0, 1, 0)) %>% 
  select(-n_strains_shared_btw_infant_pairs, -`...1`)
  
metadata.clean1 <- metadata %>% 
  select(Old_sample_name, `Mother Infant Pair`, Place_birth, `Month/Year`) %>% 
  mutate(Old_sample_name1=Old_sample_name,
         `Mother Infant Pair1`=`Mother Infant Pair`,
         Place_birth1=Place_birth,
         `Month/Year1`=`Month/Year`) %>% 
  select(-Old_sample_name, -`Mother Infant Pair`, -Place_birth, -`Month/Year`) 

metadata.clean2 <- metadata %>% 
  select(Old_sample_name, `Mother Infant Pair`, Place_birth, `Month/Year`) %>% 
  mutate(Old_sample_name2=Old_sample_name,
         `Mother Infant Pair2`=`Mother Infant Pair`,
         Place_birth2=Place_birth,
         `Month/Year2`=`Month/Year`) %>% 
  select(-Old_sample_name, -`Mother Infant Pair`, -Place_birth, -`Month/Year`) 
  
infants_loop.T2.binary.hospYear <- left_join(infants_loop.T2.binary, metadata.clean1, by=c("sample1"="Old_sample_name1"))
infants_loop.T2.binary.hospYear.extended <- left_join(infants_loop.T2.binary.hospYear, metadata.clean2, by=c("sample2"="Old_sample_name2"))

infants_loop.T2.binary.hospYear.extended <- infants_loop.T2.binary.hospYear.extended %>% 
  separate(`Month/Year1`, c("Month1", "Year1"), sep = "_", remove=FALSE) %>% 
  separate(`Month/Year2`, c("Month2", "Year2"), sep = "_", remove=FALSE)

#all unrelated infants
infants_loop.T2.binary.hospYear.extended.clean <- infants_loop.T2.binary.hospYear.extended %>% 
  filter(!is.na(Place_birth1) & !is.na(Place_birth2)) %>% 
  mutate(location=ifelse(Place_birth1==Place_birth2, "Same_Hospital", "Diff_Hospital")) %>% 
  mutate(location_time= case_when(
    (location=="Same_Hospital" & `Month/Year1`==`Month/Year2`) ~ "SameHospital-SameMY",
    (location=="Same_Hospital" & `Month/Year1`!=`Month/Year2`) ~ "SameHospital-DiffMY",
    (location=="Diff_Hospital" & `Month/Year1`==`Month/Year2`) ~ "DiffHospital-SameMY",
    (location=="Diff_Hospital" & `Month/Year1`!=`Month/Year2`) ~ "DiffHospital-DiffMY",
    TRUE ~ "Other")) %>% 
    mutate(location_timeYearOnly= case_when(
    (location=="Same_Hospital" & `Year1`==`Year2`) ~ "SameHospital-SameY",
    (location=="Same_Hospital" & `Year1`!=`Year2`) ~ "SameHospital-DiffY",
    (location=="Diff_Hospital" & `Year1`==`Year2`) ~ "DiffHospital-SameY",
    (location=="Diff_Hospital" & `Year1`!=`Year2`) ~ "DiffHospital-DiffY",
    TRUE ~ "Other"))

infants_loop.T2.binary.hospYear.extended.clean$atLeast_1strain_shared <- as.numeric(infants_loop.T2.binary.hospYear.extended.clean$atLeast_1strain_shared)
```

# Stool samples 6 months (T2) - Fisher exact tests

## Hospital Only
```{r} 
fisher.df.location.T2 <- infants_loop.T2.binary.hospYear.extended.clean %>% 
  group_by(atLeast_1strain_shared, location) %>% 
  summarise(counts=n()) %>% 
  mutate(atLeast_1strain_shared2 = ifelse(atLeast_1strain_shared==0, "No", "Yes")) %>% 
  ungroup() %>% 
  select(-atLeast_1strain_shared)

#df for fisher exact test
fisher.df.location.T2.wide <- as.data.frame(fisher.df.location.T2 %>% 
  pivot_wider(names_from = atLeast_1strain_shared2,
              values_from = counts) %>% 
  select(location,Yes, No) %>% 
  slice(2,1))

rownames(fisher.df.location.T2.wide) <- fisher.df.location.T2.wide[,1]
fisher.df.location.T2.wide[,1] <- NULL

fisher.test(fisher.df.location.T2.wide)
#Hospital Only
```

## Hospital + Year
```{r}
fisher.df.locationY.T2 <- infants_loop.T2.binary.hospYear.extended.clean %>% 
  group_by(atLeast_1strain_shared, location_timeYearOnly) %>% 
  summarise(counts=n()) %>% 
  mutate(atLeast_1strain_shared2 = ifelse(atLeast_1strain_shared==0, "No", "Yes")) %>% 
  ungroup() %>% 
  select(-atLeast_1strain_shared)

#df for fisher exact test
fisher.df.location.T2.wideY <- as.data.frame(fisher.df.locationY.T2 %>% 
  pivot_wider(names_from = atLeast_1strain_shared2,
              values_from = counts) %>% 
  select(location_timeYearOnly,Yes, No) %>% 
  slice(2,1))

rownames(fisher.df.location.T2.wideY) <- fisher.df.location.T2.wideY[,1]
fisher.df.location.T2.wideY[,1] <- NULL

fisher.test(fisher.df.location.T2.wideY)
```

# Barplot from fisher df
```{r}
#SAME HOSPITAL---------
signal.sameH_sharingYes_num.T2 <- fisher.df.location.T2.wide[1,1]
signal.sameH_sharingAll_den.T2 <- fisher.df.location.T2.wide[1,2]+fisher.df.location.T2.wide[1,1]

#baseline
baseline.sameH_sharingYes_num.T2 <- fisher.df.location.T2.wide[2,1]
baseline.sameH_sharingAll_den.T2 <- fisher.df.location.T2.wide[2,2] + fisher.df.location.T2.wide[2,1]

signal.sameH.T2 <- signal.sameH_sharingYes_num.T2/signal.sameH_sharingAll_den.T2
baseline.sameH.T2 <- baseline.sameH_sharingYes_num.T2/baseline.sameH_sharingAll_den.T2

#SAME HOSPITAL + YEAR---------
signal.sameHY_sharingYes_num.T2 <- fisher.df.location.T2.wideY[1,1]
signal.sameHY_sharingAll_den.T2 <- fisher.df.location.T2.wideY[1,2] + fisher.df.location.T2.wideY[1,1]
baseline.sameHY_sharingYes_num.T2 <- fisher.df.location.T2.wideY[2,1]
baseline.sameHY_sharingAll_den.T2 <- fisher.df.location.T2.wideY[2,2] + fisher.df.location.T2.wideY[2,1]

signal.sameHY.T2 <- signal.sameHY_sharingYes_num.T2/signal.sameHY_sharingAll_den.T2
baseline.sameHY.T2 <- baseline.sameHY_sharingYes_num.T2/baseline.sameHY_sharingAll_den.T2

Timepoint.T2 <- c("T2", "T2")
Condition.T2 <- c("SameH", "SameHY")
signal.T2 <- c(signal.sameH.T2, signal.sameHY.T2)
baseline.T2 <- c(baseline.sameH.T2, baseline.sameHY.T2)

values_df.T2 <- data.frame(Timepoint.T2, Condition.T2, signal.T2, baseline.T2)
```

## Merge T1 and T2
```{r}
values_df.T1 <- values_df.T1 %>%  mutate(Timepoint=Timepoint.T1, Condition=Condition.T1, signal=signal.T1, baseline=baseline.T1) %>% select(Timepoint, Condition, signal, baseline)
values_df.T2 <- values_df.T2 %>%  mutate(Timepoint=Timepoint.T2, Condition=Condition.T2, signal=signal.T2, baseline=baseline.T2) %>% select(Timepoint, Condition, signal, baseline)

values_df.all <- rbind(values_df.T1, values_df.T2)
values_df.all.long <- values_df.all %>% 
  pivot_longer(cols=signal:baseline, 
               names_to = "type",
               values_to = "value")

grouped.bars.plot <- ggplot(values_df.all.long, aes(fill=type, y=value, x=reorder(Condition, -value))) + 
    geom_bar(position="dodge", stat="identity", colour="black")+
    facet_grid(~Timepoint)+
    theme_bw()+
      scale_fill_manual(values=c("grey","#9ebff0"))


pdf(file=paste0("figures/strain_analysis/signal_baseline_fisherTest.pdf"), width=4.5, height=2)
grouped.bars.plot
invisible(dev.off())

grouped.bars.plot
```




