---
title: "Analysis on antimicrobial resistance genes (ARGs) in milk and infant stool samples"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)
options(stringsAsFactors = FALSE)
```

#Prepare environment.
```{r}
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(circlize)
library(gridExtra)
library("ggdendro")
library("reshape2")
library("grid")
library(magrittr)
library("ggrepel") #Avoid overlapping labels
library(data.table)
library(stringr)
library(smplot)
library(vegan)
library(devtools)
library(ComplexHeatmap)
library(dendsort)
library(smplot2)
```

#Load tables
```{r}
#DeepARG output table
arg_file <- read_delim("all_results.ARG.tsv", delim = '\t')
#metadata table
metadata <- read_delim("data/metadata.csv", delim = ',')
#predominance groups
bifido_groups <- read_delim("data/predominance_group/bifidos_groups.csv", delim = ',')
```

# PCoA on antimicrobial resistance genes
```{r}
#get sample name
names <- str_split_fixed(arg_file$filename, "_", 2)
arg_file.names <- cbind(arg_file, names)

all.samples <- c(metadata %>% 
  pull(Sample_ID))

arg_file.clean.pcoa <- arg_file.names %>% 
  filter(identity>=95) %>% 
  filter(`1` %in% all.samples)
  
arg.counts <- arg_file.clean.pcoa %>% 
  select(-`predicted_ARG-class`) %>% 
  mutate(presence=1) %>% 
  group_by(`#ARG`, `1`) %>% 
  summarise(n=n())

arg.counts.wide <- pivot_wider(arg.counts,
            names_from = `1`,
            values_from = n)

#replace NAs with 0s
arg.counts.wide[is.na(arg.counts.wide)] <- 0

arg.counts.wide_mat = arg.counts.wide |> 
  column_to_rownames("#ARG") |> 
  as.matrix() |>
  t()

dist_mat = vegdist(arg.counts.wide_mat, method = "bray")

cmd_res = cmdscale(dist_mat, 
                   k = (nrow(arg.counts.wide_mat) - 1),
                   eig = TRUE)

pcoa_df = tibble(PC1 = cmd_res$points[,1], 
                 PC2 = cmd_res$points[,2])

#Adding metadata to the plot
order.samples <- as.data.frame(a %>% 
  ungroup() %>% 
  select(`1`))

meta_ordered = left_join(order.samples, metadata, by=c(`1`="Sample_ID"))
meta_ordered.clean <- unique(meta_ordered)

pcoa_meta = bind_cols(pcoa_df, meta_ordered.clean)

pdf(file=paste0("figures/ARG_analysis/pcoa_allSampleType.pdf"), width=6, height=3.5)
ggplot(pcoa_meta, aes(x = PC1, y = PC2, color = Sample_type)) +
  geom_point(size=2.5) +
  theme_light() +
  scale_color_manual(values=c("#e66101", "#fdb863", "#5e3c99", "#b2abd2")) 
invisible(dev.off())
```

# ARGs heatmaps

## ARGs in stool samples
```{r}
stool.samples <- c(metadata %>% 
                     filter(Bodysite=="Stool") %>% 
                     pull(Old_sample_name))

arg_file.clean.stool <- arg_file.names %>% 
  filter(identity>=95) %>% 
  filter(`predicted_ARG-class`!="unclassified" & `predicted_ARG-class`!="multidrug") %>% 
  filter(`1` %in% stool.samples)
  
arg.stool.counts <- arg_file.clean.stool %>% 
  select(-`predicted_ARG-class`) %>% 
  mutate(presence=1) %>% 
  group_by(`#ARG`, `1`) %>% 
  summarise(n=n())

arg.stool.counts.wide <- pivot_wider(arg.stool.counts,
            names_from = `1`,
            values_from = n)

#replace NAs with 0s
as.data.frame(arg.stool.counts.wide[is.na(arg.stool.counts.wide)] <- 0)

arg.stool.counts.wide.mat = arg.stool.counts.wide %>% 
  column_to_rownames("#ARG") 

ARG.sums <- as.data.frame(rowSums(arg.stool.counts.wide.mat)) 

ARG_keep <- c(ARG.sums %>% 
  rownames_to_column(var="ARG") %>% 
  filter(`rowSums(arg.stool.counts.wide.mat)`>200) %>% 
  pull(ARG))

arg.stool.counts.wide.mat.clean <- arg.stool.counts.wide.mat %>% 
  rownames_to_column(var="ARG") %>% 
  filter(ARG %in% ARG_keep) %>% 
  column_to_rownames("ARG") %>% 
  as.matrix()

#get sample names that will be in heatmap
samples.heatmap <- colnames(arg.stool.counts.wide.mat.clean)

metadata.bifidos <- left_join(metadata, bifido_groups, by=c("Full_name"="Sample"))
metadata.bifidos.clean <- metadata.bifidos %>% 
  mutate(Delivery_mode_vag=ifelse(Delivery_mode_cat1=="AVD" | Delivery_mode_cat1=="SVD", "Yes", "No")) %>% 
  filter(Old_sample_name %in% samples.heatmap) %>%  #extract metadata from only rows that will be in heatmap 
  arrange(factor(Old_sample_name, levels = samples.heatmap))

#*****ANNOTATION COLUMNS 
ann <- data.frame(metadata.bifidos.clean$`Sample_type`,
                  metadata.bifidos.clean$`Delivery_mode_vag`,
                  metadata.bifidos.clean$`Exclusive_bf`,
                 metadata.bifidos.clean$`Abx_exposure_any`,
                 metadata.bifidos.clean$`Group`)

colnames(ann) <- c('Sample_type', 'Delivery_mode_vag','Exclusive_bf', 'Abx_exposure_any', 'Group')

colours <- list('Sample_type' = c('Stool_1M' = "#5e3c99", 'Stool_6M' = "#b2abd2"),
  'Delivery_mode_vag' = c('Yes' = 'black', 'No' = 'cornflowerblue'),
  'Exclusive_bf' = c('Yes' = 'black', 'No' = 'cornflowerblue'),
  'Abx_exposure_any' = c('Yes' = 'black', 'No' = 'cornflowerblue'),
  'Group' = c('Group_1' = '#630a4c', 'Group_2' = '#ad3691', 'Group_3' = '#de9ecd', 'Group_4' = '#f2e6ef'))

colAnn <- HeatmapAnnotation(df = ann,
  which = 'col',
  col = colours,
  annotation_width = unit(c(0.1, 0.1), 'cm'),
  gap = unit(0.5, 'mm'))

#*****ANNOTATION ROWS (ARG names)
#arg.heatmap <- c(rownames(b_mat_heatmap.clean)) #get agr names that will be in heatmap
arg.heatmap <- as.data.frame(arg.stool.counts.wide.mat.clean) %>% 
  rownames_to_column(var="name") %>% 
  select(name)

arg_file.names.small <- arg_file.names %>% 
  select("#ARG", "predicted_ARG-class") %>% 
  unique()

b_mat_heatmap.clean.ARGnames.matadata <- left_join(arg.heatmap, arg_file.names.small, by=c("name"="#ARG"))

#check corner case: genes that fall into multiple ARG classes
#here: MDTK multidrug + fluoroquinolone -> chosen fluoroquinolone
#and: ARNA polymyxin + peptide -> chosen peptide
counts <- b_mat_heatmap.clean.ARGnames.matadata %>% 
  group_by(name) %>% 
  summarise(n=n())

#get top n classes and collapse the others under the "other" label
top_classes <- c(b_mat_heatmap.clean.ARGnames.matadata %>% 
  group_by(`predicted_ARG-class`) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n)) %>% 
  top_n(5) %>% 
  pull(`predicted_ARG-class`))
  
b_mat_heatmap.clean.ARGnames.matadata.noMulticlass <- arg.stool.counts.wide.mat.clean.ARGnames.matadata %>% 
  filter(!((name=="MDTK" & `predicted_ARG-class` == "multidrug") | (name=="ARNA" & `predicted_ARG-class` == "polymyxin"))) %>% 
  mutate(class_clean = ifelse(`predicted_ARG-class` %in% top_classes, `predicted_ARG-class`, "other"))
  
classes <- as.data.frame(b_mat_heatmap.clean.ARGnames.matadata.noMulticlass) %>% 
  group_by(`predicted_ARG-class`) %>% 
  summarise(n=n())

ann2 <- data.frame(b_mat_heatmap.clean.ARGnames.matadata.noMulticlass$class_clean)
colnames(ann2) <- c('class_clean')

rowcolours <- list(
  'class_clean' = c('beta-lactam' = 'darkred', 'tetracycline' = 'darkorange',
                    'fluoroquinolone' = 'gold', 'MLS' = 'navy', 
                    'aminoglycoside' = 'cornflowerblue', 'other' = 'white', 'diaminopyrimidine'="black"))

rowAnn <- rowAnnotation(df = ann2,
  col = rowcolours,
  annotation_width = unit(c(0.1, 0.1), 'cm'),
  gap = unit(0.5, 'mm'))
#end annotations

row_dend = dendsort(hclust(dist(arg.stool.counts.wide.mat.clean)))
col_dend = dendsort(hclust(dist(t(arg.stool.counts.wide.mat.clean))))

col_fun = colorRamp2(c(0, 1, 10, 100, 200, 500), 
                     c("white","#e5e8f2","#ccd1e6", "#6776b5", '#35499d', "#000000"))

pdf(file=paste0("figures/ARG_analysis/heatmap_stool.pdf"), width=10, height=8)
Heatmap(arg.stool.counts.wide.mat.clean, 
        name = "Occurrences", 
        #cluster_columns = col_dend,
        clustering_distance_rows = "euclidean", 
        clustering_distance_columns = "euclidean",
        show_column_names = FALSE,
        show_row_names = FALSE,
        show_row_dend = TRUE, 
        col = col_fun, 
        na_col = "red",
        row_names_gp = gpar(fontsize = 8),
        top_annotation=colAnn,
        left_annotation = rowAnn,
        column_dend_height = unit(5, "mm"))
invisible(dev.off())
```

## ARGs in milk samples
```{r}
milk.samples <- c(metadata %>% 
                     filter(Bodysite=="Milk") %>% 
                     pull(Old_sample_name))

arg_file.clean.heatmap <- arg_file.names %>% 
  filter(identity>=95) %>% 
  filter(`predicted_ARG-class`!="unclassified" & `predicted_ARG-class`!="multidrug") %>% 
  filter(`1` %in% milk.samples)
  
arg.milk.counts <- arg_file.clean.heatmap %>% 
  select(-`predicted_ARG-class`) %>% 
  mutate(presence=1) %>% 
  group_by(`#ARG`, `1`) %>% 
  summarise(n=n())

arg.milk.counts.wide <- pivot_wider(arg.milk.counts,
            names_from = `1`,
            values_from = n)

#replace NAs with 0s
as.data.frame(arg.milk.counts.wide[is.na(arg.milk.counts.wide)] <- 0)

arg.milk.counts.wide.mat = arg.milk.counts.wide %>% 
  column_to_rownames("#ARG") 

ARG.sums <- as.data.frame(rowSums(arg.milk.counts.wide.mat)) 

ARG_keep <- c(ARG.sums %>% 
  rownames_to_column(var="ARG") %>% 
  filter(`rowSums(arg.milk.counts.wide.mat)`>10) %>% 
  pull(ARG))

arg.milk.counts.wide.mat.clean <- arg.milk.counts.wide.mat %>% 
  rownames_to_column(var="ARG") %>% 
  filter(ARG %in% ARG_keep) %>% 
  column_to_rownames("ARG") %>% 
  as.matrix()

#get sample names that will be in heatmap
samples.heatmap <- colnames(arg.milk.counts.wide.mat.clean)

metadata.bifidos <- left_join(metadata, bifido_groups, by=c("Full_name"="Sample"))
metadata.bifidos.clean <- metadata.bifidos %>% 
  filter(Old_sample_name %in% samples.heatmap) %>%  #extract metadata from only rows that will be in heatmap 
  arrange(factor(Old_sample_name, levels = samples.heatmap))

#*****ANNOTATION COLUMNS 
ann <- data.frame(metadata.bifidos.clean$`Sample_type`,
                  metadata.bifidos.clean$`Status`,
                  metadata.bifidos.clean$`Group`)

colnames(ann) <- c('Sample_type', 'Status', 'Group')

colours <- list('Sample_type' = c('Milk_1M' = "#e66101", 'Milk_3M' = "#fdb863"),
  'Status' = c('GD' = 'black', 'Healthy' = 'cornflowerblue'),
  'Group' = c('Group_1' = '#630a4c', 'Group_2' = '#ad3691', 'Group_3' = '#de9ecd', 'Group_4' = '#f2e6ef'))

colAnn <- HeatmapAnnotation(df = ann,
  which = 'col',
  col = colours,
  annotation_width = unit(c(0.1, 0.1), 'cm'),
  gap = unit(0.5, 'mm'))

#*****ANNOTATION ROWS (ARG names)
arg.heatmap <- as.data.frame(arg.milk.counts.wide.mat.clean) %>% 
  rownames_to_column(var="name") %>% 
  select(name)

arg_file.names.small <- arg_file.names %>% 
  select("#ARG", "predicted_ARG-class") %>% 
  unique()

b_mat_heatmap.clean.ARGnames.matadata <- left_join(arg.heatmap, arg_file.names.small, by=c("name"="#ARG"))

#check corner case: genes that fall into multiple ARG classes
#here: MDTK multidrug + fluoroquinolone -> chosen fluoroquinolone
#and: ARNA polymyxin + peptide -> chosen peptide
counts <- arg.milk.counts.wide.mat.clean.ARGnames.matadata %>% 
  group_by(name) %>% 
  summarise(n=n())

#get top n classes and collapse the others under the "other" label
top_classes <- c(arg.milk.counts.wide.mat.clean.ARGnames.matadata %>% 
  group_by(`predicted_ARG-class`) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n)) %>% 
  top_n(5) %>% 
  pull(`predicted_ARG-class`))
  
arg.milk.counts.wide.mat.clean.ARGnames.matadata.noMulticlass <- 
  arg.milk.counts.wide.mat.clean.ARGnames.matadata %>% 
  filter(!((name=="MDTK" & `predicted_ARG-class` == "multidrug") | (name=="ARNA" & `predicted_ARG-class` == "polymyxin"))) %>% 
  mutate(class_clean = ifelse(`predicted_ARG-class` %in% top_classes, `predicted_ARG-class`, "other"))
  
classes <- as.data.frame(arg.milk.counts.wide.mat.clean.ARGnames.matadata.noMulticlass) %>% 
  group_by(`predicted_ARG-class`) %>% 
  summarise(n=n())

ann2 <- data.frame(arg.milk.counts.wide.mat.clean.ARGnames.matadata.noMulticlass$class_clean)
colnames(ann2) <- c('class_clean')

rowcolours <- list(
  'class_clean' = c('beta-lactam' = 'darkred', 'tetracycline' = 'darkorange',
                    'peptide' = 'gold', 'MLS' = 'navy', 
                    'aminoglycoside' = 'cornflowerblue', 'other' = 'white'))

rowAnn <- rowAnnotation(df = ann2,
  col = rowcolours,
  annotation_width = unit(c(0.1, 0.1), 'cm'),
  gap = unit(0.5, 'mm'))
#end annotations

row_dend = dendsort(hclust(dist(arg.milk.counts.wide.mat.clean)))
col_dend = dendsort(hclust(dist(t(arg.milk.counts.wide.mat.clean))))

col_fun = colorRamp2(c(0, 1, 10, 100, 200, 500), 
                     c("white","#e5e8f2","#ccd1e6", "#6776b5", '#35499d', "#000000"))

pdf(file=paste0("figures/ARG_analysis/heatmap_milk.pdf"), width=10, height=8)
Heatmap(arg.milk.counts.wide.mat.clean, 
        name = "Occurrences", 
        clustering_distance_rows = "euclidean", 
        clustering_distance_columns = "euclidean",
        show_column_names = FALSE,
        show_row_names = FALSE,
        show_row_dend = TRUE, 
        col = col_fun, 
        na_col = "red",
        row_names_gp = gpar(fontsize = 8),
        top_annotation=colAnn,
        left_annotation = rowAnn,
        column_dend_height = unit(5, "mm"))
invisible(dev.off())
```

# Identity threshold for ARGs
```{r}
#identity thresholds
arg_file.clean <- arg_file.names %>% 
  filter(identity>=95) %>% 
  select(`1`, `#ARG`, `predicted_ARG-class`) %>% 
  filter(`predicted_ARG-class` != "multidrug")

arg_file.clean_metadata <- left_join(arg_file.clean, metadata, by=c("1"= "Correct_sample_name"))
  
arg_file.clean_metadata <- arg_file.clean_metadata %>% 
  filter(Post_filtering_QC=="Keep", `predicted_ARG-class` != "unclassified") %>% 
  select(-`1`, -Reads_filtered, -Reads_mapped, -Percent, -`Total taxa`, -Post_filtering_QC)
```


# ARGs prevalences statified by relevant variables:

## by predominance group
```{r}
arg.master <- arg_file.names %>% 
  filter(identity>=95) %>% 
  filter(`predicted_ARG-class`!="unclassified" & `predicted_ARG-class`!="multidrug") 

arg.master.groups <- left_join(arg.master, metadata.bifidos %>% select(Old_sample_name,Group, Sample_type), by=c('1'="Old_sample_name")) %>% 
    filter(Sample_type=="Stool_1M" | Sample_type=="Stool_6M") %>% 
    mutate(Bifido_group=ifelse(Group=="Group_4", "No", "Yes")) #combine all bifido groups vs not

predom.group.master <- arg.master.groups %>% 
  filter(!is.na(Bifido_group)) %>% 
  group_by(Bifido_group, `1`, Sample_type) %>% 
  summarise(counts=n())

pdf(file=paste0("figures/ARG_analysis/arg_bifido_groups_boxplots.pdf"), width=4, height=2)
ggplot(predom.group.master, aes(fill=Bifido_group, y=counts, x=Sample_type)) + 
  geom_boxplot(outlier.size = -1) +
  #stat_compare_means(method = "t.test") +
  stat_compare_means(method = "t.test", label = "p.signif") +
  scale_fill_manual(values=c('#630a4c', '#ad3691', '#de9ecd', '#f2e6ef'))+
  theme_bw()
invisible(dev.off())
```

## by delivery mode
```{r}
arg_file.names <- cbind(arg_file, names)

arg.master <- arg_file.names %>% 
  filter(identity>=95) %>% 
  filter(`predicted_ARG-class`!="unclassified" & `predicted_ARG-class`!="multidrug")

metadata.bifidos.clean.vag <- metadata.bifidos %>% 
  mutate(Delivery_mode_vag=ifelse(Delivery_mode_cat1=="AVD" | Delivery_mode_cat1=="SVD", "Yes", "No"))  

arg.master.vag <- left_join(arg.master, metadata.bifidos.clean.vag %>% select(Old_sample_name,Group, Sample_type, Delivery_mode_vag), by=c('1'="Old_sample_name"))

partial1.vag <- arg.master.vag %>% 
  mutate(class_clean=ifelse(`predicted_ARG-class` %in% top.classes, `predicted_ARG-class`, "other")) %>% 
  filter(!is.na(Delivery_mode_vag))%>% 
  group_by(Sample_type, Delivery_mode_vag) %>% 
  summarise(n=n()) %>% 
  select(Sample_type, Delivery_mode_vag, n)

partial2.vag <- arg.master.vag %>% 
  filter(!is.na(Delivery_mode_vag)) %>% 
  mutate(class_clean=ifelse(`predicted_ARG-class` %in% top.classes, `predicted_ARG-class`, "other")) %>% 
  group_by(Sample_type) %>% 
  summarise(totals=n())

merged.vag <- left_join(partial2.vag, partial1.vag, by="Sample_type")

merged.vag <- merged.vag %>% 
  mutate(tmp=paste0(Sample_type,"_",Delivery_mode_vag)) 

nsamples <- metadata.bifidos.clean.vag %>% 
  group_by(Sample_type, Delivery_mode_vag) %>% 
  summarise(nsamples=n()) %>% 
  mutate(tmp=paste0(Sample_type,"_",Delivery_mode_vag)) %>% 
  select(-Sample_type, -Delivery_mode_vag)

merged.vag.nsamples <- left_join(merged.vag, nsamples, by="tmp")

merged.vag.nsamples <- merged.vag.nsamples %>% 
  mutate(perc=n*100/totals, ratio=perc/nsamples) %>% 
  mutate(argnsamples=n/nsamples) %>% 
  filter(Sample_type.x=="Stool_1M"|Sample_type.x=="Stool_6M")

#boxplot
arg.perSample.vag <- arg.master.vag %>% 
  mutate(Sample=`1`) %>% 
  filter(!is.na(Delivery_mode_vag))%>% 
  filter(Sample_type=="Stool_1M" | Sample_type=="Stool_6M") %>% 
  group_by(Sample) %>% 
  summarise(nARG_perSample=n()) 

arg.master.vag.meta <- arg.master.vag %>% 
  mutate(Sample=`1`) %>% 
  select(Sample, Sample_type, Delivery_mode_vag, Group) 
  
arg.perSample.vag.meta <- left_join(arg.perSample.vag, arg.master.vag.meta, by="Sample")
arg.perSample.vag.meta <- arg.perSample.vag.meta %>% distinct()

stat_box_data <- function(y) {
  return(data.frame(y = 0, label = paste(length(y))))
}

pdf(file=paste0("figures/ARG_analysis/arg_vag_boxplot.pdf"), width=5, height=2.5)
ggplot(arg.perSample.vag.meta, aes(x=Sample_type, y=nARG_perSample, fill=Delivery_mode_vag)) +
  geom_boxplot(outlier.shape=NA, position=position_dodge()) +
  theme_bw() +
  scale_fill_manual(values=c("#5e3c99", "#b2abd2")) +
  coord_flip()+
  stat_compare_means(aes(group = Delivery_mode_vag), method = "t.test") +
  #stat_compare_means(aes(group = Delivery_mode_vag), method = "t.test", label = "p.signif") +
  stat_summary(fun.data = stat_box_data, geom = "text", fun = median, position = position_dodge(width = 0.85), size=2)
invisible(dev.off())

#pvalues (adjusted)
anno_df.vag = compare_means(nARG_perSample ~ Delivery_mode_vag, group.by = "Sample_type", data = arg.perSample.vag.meta, method="t.test")

```

## by sample type
```{r}
arg_file.names <- cbind(arg_file, names)

arg.master <- arg_file.names %>% 
  filter(identity>=95) %>% 
  filter(`predicted_ARG-class`!="unclassified" & `predicted_ARG-class`!="multidrug")

arg.master.metadata <- left_join(arg.master, metadata %>% select(Old_sample_name, Sample_type), by=c('1'="Old_sample_name"))

#number of genes (decided to drop this plot)
nARG_sampleType <- arg.master.metadata %>% 
  group_by(`1`, Sample_type) %>%
  summarise(nARG=n())

#***** number of distinct ARG classes
nARG_classes_sampleType <- arg.master.metadata %>% 
  select(`1`, `predicted_ARG-class`, Sample_type) %>% 
  unique() %>% 
  group_by(`1`, Sample_type) %>% 
  summarise(nARG_classes=n())

my_comparisons <- list( c("Milk_1M", "Milk_3M"), c("Stool_1M", "Stool_6M"), c("Milk_3M", "Stool_1M") )

pdf(file=paste0("figures/ARG_analysis/arg_classes_sampleType_boxplot.pdf"), width=3, height=3)
ggplot(nARG_classes_sampleType, aes(x=Sample_type, y=nARG_classes, fill=Sample_type)) +
  geom_boxplot(outlier.shape=NA, position=position_dodge()) +
  theme_bw() +
  scale_fill_manual(values=c("#e66101", "#fdb863", "#5e3c99", "#b2abd2")) +
  stat_compare_means(comparisons = my_comparisons, label = "p.signif", method = "t.test") 
invisible(dev.off())
```

## by antibiotic intake
```{r}
arg_file.names <- cbind(arg_file, names)

arg.master <- arg_file.names %>% 
  filter(identity>=95) %>% 
  filter(`predicted_ARG-class`!="unclassified" & `predicted_ARG-class`!="multidrug")

arg.master.abx <- left_join(arg.master, metadata.bifidos %>% select(Old_sample_name,Group, Sample_type, Abx_exposure_any), by=c('1'="Old_sample_name"))

partial1.abx <- arg.master.abx %>% 
  mutate(class_clean=ifelse(`predicted_ARG-class` %in% top.classes, `predicted_ARG-class`, "other")) %>% 
  filter(!is.na(Abx_exposure_any))%>% 
  group_by(Sample_type, Abx_exposure_any) %>% 
  summarise(n=n()) %>% 
  select(Sample_type, Abx_exposure_any, n)

partial2.abx <- arg.master.abx %>% 
  filter(!is.na(Abx_exposure_any)) %>% 
  mutate(class_clean=ifelse(`predicted_ARG-class` %in% top.classes, `predicted_ARG-class`, "other")) %>% 
  group_by(Sample_type) %>% 
  summarise(totals=n())

merged.abx <- left_join(partial2.abx, partial1.abx, by="Sample_type")

merged.abx <- merged.abx %>% 
  mutate(tmp=paste0(Sample_type,"_",Abx_exposure_any)) 

nsamples <- metadata.bifidos %>% 
  group_by(Sample_type, Abx_exposure_any) %>% 
  summarise(nsamples=n()) %>% 
  mutate(tmp=paste0(Sample_type,"_",Abx_exposure_any)) %>% 
  select(-Sample_type, -Abx_exposure_any)

merged.abx.nsamples <- left_join(merged.abx, nsamples, by="tmp")

merged.abx.nsamples <- merged.abx.nsamples %>% 
  mutate(perc=n*100/totals, ratio=perc/nsamples) %>% 
  mutate(argnsamples=n/nsamples) %>% 
  filter(Sample_type.x=="Stool_1M"|Sample_type.x=="Stool_6M")

#boxplot
arg.perSample.abx <- arg.master.abx %>% 
  mutate(Sample=`1`) %>% 
  filter(!is.na(Abx_exposure_any))%>% 
  filter(Sample_type=="Stool_1M" | Sample_type=="Stool_6M") %>% 
  group_by(Sample) %>% 
  summarise(nARG_perSample=n()) 

arg.master.abx.meta <- arg.master.abx %>% 
  mutate(Sample=`1`) %>% 
  select(Sample, Sample_type, Abx_exposure_any, Group) 
  
arg.perSample.abx.meta <- left_join(arg.perSample.abx, arg.master.abx.meta, by="Sample")
arg.perSample.abx.meta <- arg.perSample.abx.meta %>% distinct()

pdf(file=paste0("figures/ARG_analysis/arg_abx_boxplot.pdf"), width=5, height=2.5)
ggplot(arg.perSample.abx.meta, aes(x=Sample_type, y=nARG_perSample, fill=Abx_exposure_any)) +
  geom_boxplot(outlier.shape=NA, position=position_dodge()) +
  theme_bw() +
  scale_fill_manual(values=c("#5e3c99", "#b2abd2")) +
  coord_flip()+
  stat_compare_means(aes(group = Abx_exposure_any), label = "p.signif", method = "t.test") +
  stat_summary(fun.data = stat_box_data, geom = "text", fun = median, position = position_dodge(width = 0.85), size=2)
invisible(dev.off())

#pvalues (adjusted)
anno_df.abx = compare_means(nARG_perSample ~ Abx_exposure_any, group.by = "Sample_type", data = arg.perSample.abx.meta)
```


# ARGs carriage overview across sample types
```{r}
occurrences <- arg_file.clean_metadata %>% 
  group_by(Sample_type, `predicted_ARG-class`) %>% 
  summarise(n=n()) 

totals <- arg_file.clean_metadata %>% 
  group_by(Sample_type) %>% 
  summarise(tot=n())

arg.merged <- left_join(occurrences, totals, by=c("Sample_type"))
  
arg.merged.perc <- arg.merged %>%
  mutate(perc=n*100/tot)

#MILK 1M
arg.merged.perc.M1M.classes <- arg.merged.perc %>%
  filter(Sample_type=="Milk_1M" & perc>3) %>% 
  select(Sample_type, `predicted_ARG-class`, perc)

valueM1M <- as.double(arg.merged.perc %>% 
  filter(Sample_type=="Milk_1M") %>% 
  filter(perc<3) %>% 
  summarise(sum=sum(perc)) %>% 
  select(sum) %>% pull(sum))

df2<-data.frame("Milk_1M","Other", valueM1M)
names(df2)<-c("Sample_type","predicted_ARG-class", "perc")

M1M <- rbind(arg.merged.perc.M1M.classes, df2)

#MILK 3M
arg.merged.perc.M3M.classes <- arg.merged.perc %>%
  filter(Sample_type=="Milk_3M" & perc>3) %>% 
  select(Sample_type, `predicted_ARG-class`, perc)

valueM3M <- as.double(arg.merged.perc %>% 
  filter(Sample_type=="Milk_3M") %>% 
  filter(perc<3) %>% 
  summarise(sum=sum(perc)) %>% 
  select(sum) %>% pull(sum))

df3<-data.frame("Milk_3M","Other", valueM3M)
names(df3)<-c("Sample_type","predicted_ARG-class", "perc")

M3M <- rbind(arg.merged.perc.M3M.classes, df3)

#STOOL 1M
arg.merged.perc.S1M.classes <- arg.merged.perc %>%
  filter(Sample_type=="Stool_1M" & perc>3) %>% 
  select(Sample_type, `predicted_ARG-class`, perc)

valueS1M <- as.double(arg.merged.perc %>% 
  filter(Sample_type=="Stool_1M") %>% 
  filter(perc<3) %>% 
  summarise(sum=sum(perc)) %>% 
  select(sum) %>% pull(sum))

df4<-data.frame("Stool_1M","Other", valueS1M)
names(df4)<-c("Sample_type","predicted_ARG-class", "perc")

S1M <- rbind(arg.merged.perc.S1M.classes, df4)

#STOOL 6M
arg.merged.perc.S6M.classes <- arg.merged.perc %>%
  filter(Sample_type=="Stool_6M" & perc>3) %>% 
  select(Sample_type, `predicted_ARG-class`, perc)

valueS6M <- as.double(arg.merged.perc %>% 
  filter(Sample_type=="Stool_6M") %>% 
  filter(perc<3) %>% 
  summarise(sum=sum(perc)) %>% 
  select(sum) %>% pull(sum))

df5<-data.frame("Stool_6M","Other", valueS6M)
names(df5)<-c("Sample_type","predicted_ARG-class", "perc")

S6M <- rbind(arg.merged.perc.S6M.classes, df5)

arg.sampleTypes.all <- rbind(M1M, M3M, S1M, S6M)
arg.sampleTypes.all$perc <- as.numeric(arg.sampleTypes.all$perc)

M1M.tmp <- M1M %>% 
  mutate_if(is.numeric, ~round(., 0)) %>% 
  group_by(Sample_type) %>% 
  mutate(sum=sum(perc))

M3M.tmp <- M3M %>% 
  mutate_if(is.numeric, ~round(., 0)) %>% 
  group_by(Sample_type) %>% 
  mutate(sum=sum(perc))

S1M.tmp <- S1M %>% 
  mutate_if(is.numeric, ~round(., 0)) %>% 
  group_by(Sample_type) %>% 
  mutate(sum=sum(perc))

S6M.tmp <- S6M %>% 
  mutate_if(is.numeric, ~round(., 0)) %>% 
  group_by(Sample_type) %>% 
  mutate(sum=sum(perc))

pdf(file=paste0("figures/ARG_analysis/arg_overview.pdf"), width=6, height=2.5)
  ggplot(arg.sampleTypes.all.rounded, aes(x = Sample_type, y = perc, fill = `predicted_ARG-class`, label=perc)) + 
    geom_bar(stat = "identity", colour="white")+
    theme_bw()+
    coord_flip()+
    geom_text(size = 3, position = position_stack(vjust = 0.5))+
    scale_fill_manual(values = c("#D45500", "#FF7F2A", "#FFB280", "#FFCC00","#89A02C","#BCD35F","#87DECD","#D5E5FF","#80B3FF","#3671C8","#162D50"))
invisible(dev.off())
```

# Correlation between ARGs in milk vs stool or stools 1M vs 6M

## Milk samples vs stool samples
```{r}
bothBodysites <- arg_file.clean_metadata %>% 
  select(Bodysite, `Mother Infant Pair`, `predicted_ARG-class`) %>% 
  group_by(Bodysite, `Mother Infant Pair`, `predicted_ARG-class`) %>% 
  distinct() %>% 
  group_by(`Mother Infant Pair`, `predicted_ARG-class`) %>% 
  summarise(n=n()) %>% 
  filter(n>1)

bothBodysites_couple <- c(bothBodysites %>% 
  pull(`Mother Infant Pair`))

bothBodysites_class <- c(bothBodysites %>% 
  pull(`predicted_ARG-class`))

bothBodysites_class_top <- arg_file.clean_metadata %>% 
  group_by(`predicted_ARG-class`) %>% 
  summarise(n=n()) %>% 
  top_n(5) %>% 
  pull(`predicted_ARG-class`)

arg.classes.couple <- arg_file.clean_metadata %>% 
  filter((`Mother Infant Pair` %in% bothBodysites_couple) &
           (`predicted_ARG-class` %in% bothBodysites_class) &
           (`predicted_ARG-class` %in% bothBodysites_class_top)) %>% 
  #filter(`Mother Infant Pair`!= "Mother_Infant_Pair_1" & `Mother Infant Pair`!= "Mother_Infant_Pair_3") %>%
  group_by(Bodysite, `Mother Infant Pair`, `predicted_ARG-class`) %>% 
  summarise(n=n()) 

arg.classes.couple.wide <- arg.classes.couple %>% 
  pivot_wider(names_from = "Bodysite", 
              values_from = "n")

corr1 <- ggplot(data = arg.classes.couple.wide, mapping = aes(x = Stool, y = Milk)) +
  geom_point(shape = 21, fill = 'grey55', alpha=0.5, color = 'black', size = 1.8) +
  theme_bw() +
  sm_statCorr(color = 'black', corr_method = 'spearman', linetype = 'dashed')
```

## Stool samples at 1 month vs those at 6 months of age
```{r}
bothTimepoints <- arg_file.clean_metadata %>% 
  filter(Bodysite=="Stool") %>% 
  select(Sample_type, `Mother Infant Pair`, `predicted_ARG-class`) %>% 
  group_by(Sample_type, `Mother Infant Pair`, `predicted_ARG-class`) %>% 
  distinct() %>% 
  group_by(`Mother Infant Pair`, `predicted_ARG-class`) %>% 
  summarise(n=n()) %>% 
  filter(n>1)

bothTimepoints_longitudinal <- c(bothTimepoints %>% 
  pull(`Mother Infant Pair`))

bothTimepoints_class <- c(bothTimepoints %>% 
  pull(`predicted_ARG-class`))

bothTimepoints_class_top <- arg_file.clean_metadata %>% 
  filter(Bodysite=="Stool") %>% 
  group_by(`predicted_ARG-class`) %>% 
  summarise(n=n()) %>% 
  top_n(15) %>% 
  pull(`predicted_ARG-class`)

arg.classes.longitudinal <- arg_file.clean_metadata %>% 
  filter(Bodysite=="Stool") %>% 
     filter((`Mother Infant Pair` %in% bothTimepoints_longitudinal) & 
            (`predicted_ARG-class` %in% bothTimepoints_class) &
            (`predicted_ARG-class` %in% bothTimepoints_class_top)) %>% 
  group_by(Sample_type, `Mother Infant Pair`, `predicted_ARG-class`) %>% 
  summarise(n=n()) 

arg.classes.longitudinal.wide <- arg.classes.longitudinal %>% 
  pivot_wider(names_from = "Sample_type", 
              values_from = "n")

corr2 <- ggplot(data = arg.classes.longitudinal.wide, mapping = aes(x = Stool_1M, y = Stool_6M)) +
  geom_point(shape = 21, fill = 'grey55', alpha=0.5, color = 'black', size = 1.8) +
  theme_bw() +
  sm_statCorr(color = 'black', corr_method = 'spearman', linetype = 'dashed')

#get exact pvalues with cor.test()
res <- cor.test(arg.classes.longitudinal.wide$Stool_1M, arg.classes.longitudinal.wide$Stool_6M, 
                    method = "spearman")
res$p.value
res$estimate

#combine plots
pdf(file=paste0("figures/ARG_analysis/corr_all.pdf"), width=5, height=2.3)
ggarrange(corr1, corr2, labels = c("C", "D"), widths = c(1,1.1), align = "h", nrow = 1)
invisible(dev.off())
```


