---
title: "Analaysis of species stability over time and differential abundance between exclusive breastfed and not exclusively breastfed infants"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)
options(stringsAsFactors = FALSE)
```

## Prepare environment
```{r}
library(tidyverse)
library(ggpubr)
library(ggplot2)
library(gridExtra)
library("grid")
library(magrittr)
library(data.table)
library(stringr)
library(viridis)
library(hrbrthemes)
library(ComplexHeatmap)
```

## Load tables
```{r}
#metaphlan output table
mpa_file <- read_delim("data/species_composition/milk_infant_rel_abundance.txt", delim = '\t')
metadata <- read_delim("data/metadata.csv", delim = ',')
```

## Extract species information
```{r}
#separate full clade name into taxonomic levels
mpa_file.delim <- mpa_file %>% 
  tidyr::separate(clade_name, c('Kingdom', 'Phylum', 'Class', 'Order', 'Family', 'Genus', 'Species', 'Strain'), sep ='\\|', remove = FALSE, fill="right") 

#select clades assigned to species level
mpa_file.clean.species <- mpa_file.delim %>% 
  filter(!is.na(Species) & is.na(Strain)) 

#keep simplified species name only
to.drop <- c("clade_name", "Kingdom", "Phylum", "Class", "Order", "Family", "Genus", "Strain")
mpa_file.species = mpa_file.clean.species[,!(names(mpa_file.clean.species) %in% to.drop)]
mpa_file.species$Species <- str_remove(mpa_file.species$Species, pattern = "s__")
mpa_file.species$Species <- gsub("_", " ", mpa_file.species$Species)

mpa_file.species.long <- mpa_file.species %>% 
  pivot_longer(cols = !Species, names_to = "Sample", values_to = "Abundance") 

#split sample ID for later merge with metadata table
mpa_file.species.long <- mpa_file.species.long %>% 
  mutate(full_ID=Sample) %>% 
  tidyr::separate(Sample, c('Sample', 'number'), sep ='_', remove = FALSE, fill="right") 

#merge with metadata table
merged.species <- left_join(mpa_file.species.long, metadata, by=c("full_ID"="Sample_ID"))
merged.species.clean <- merged.species %>% 
  select(Species, Sample, Abundance, Sample_type, `Mother Infant Pair`, Exclusive_bf) %>% 
  filter(Abundance>0) %>% 
  distinct() %>% 
  filter(Sample_type=="Stool_1M" | Sample_type=="Stool_6M")
```


# Species persistence over time in infant stool samples

## filtering criteria for mother-infant pairs and most prevalent species
```{r}
#get species prevalence across 1M and 6M infant stools
species.prevalence <- merged.species.clean %>%
  group_by(Species) %>%
  mutate(n_tmp=n()) %>% 
  select(Species, n_tmp) %>% 
  distinct() 

#get top 10 most prevalent species
top10.species <- species.prevalence[order(desc(species.prevalence$n_tmp)),] %>% 
  head(n=15) %>% 
  pull(Species)

#get stool samples only from timeseries for which we have both timepoints
couples_bothStool <- c(merged.species.clean %>% 
  select(`Mother Infant Pair`, Sample_type) %>% 
  distinct()%>% 
  group_by(`Mother Infant Pair`) %>% 
  summarise(n=n()) %>% 
  filter(n==2) %>% 
  pull(`Mother Infant Pair`))
```

## Overview of species stability across all infants
```{r}
merged.species.clean.couplesSpecies <- merged.species.clean %>% 
  filter((`Mother Infant Pair` %in% couples_bothStool) & (Species %in% top10.species)) %>% 
  mutate(Timepoint = ifelse(Sample_type=="Stool_1M", "1", "6")) %>% 
  group_by(Timepoint, Species) %>% 
  summarise(mean_rel_abs=mean(Abundance)) %>% 
  group_by(Timepoint) %>% 
  mutate(total_rel_abs = sum(mean_rel_abs),
         percentage = mean_rel_abs*100/total_rel_abs)

custom.stool3 <- c("#E57373", "#EC407A", "#D816B0", "#AB47BC", "#8E24AA", "#7E57C2", "#5E35B1", "#5C6BC0", "#3949AB", "#1A237E", "black", "white", "red", "blue", "green")

merged.species.clean.couplesSpecies <- merged.species.clean.couplesSpecies %>% 
  select(Species, Timepoint, percentage)

merged.species.clean.couplesSpecies$Timepoint <- as.numeric(merged.species.clean.couplesSpecies$Timepoint)

all.stool.stability.plot <- ggplot(merged.species.clean.couplesSpecies, aes(x=Timepoint, y=percentage, fill=Species)) + 
  geom_area(position = 'stack', alpha=0.9, size=.5, colour="black")+
  theme_light() +
  scale_fill_manual(values=rev(custom.stool3)) 

pdf(file=paste0("figures/species_stability/overview_stool_stability.pdf"), width=5, height=4)
all.stool.stability.plot
invisible(dev.off())

all.stool.stability.plot
```

## t-test differentially abundant 1M vs 6M
```{r}

tmp1 <- merged.species.clean %>% 
  mutate(Timepoint = ifelse(Sample_type=="Stool_1M", "1", "6")) %>% 
  filter((`Mother Infant Pair` %in% couples_bothStool) & (Species %in% top10.species)) 
  
collect.results.allInfants <- data.frame()

#species in both timepoints
species.list.bothTimepoints.allInfants <- c(tmp1 %>% 
  group_by(Species, `Mother Infant Pair`) %>% 
  summarise(n=n()) %>% 
  filter(n==2) %>% 
  pull(Species) %>% 
  unique())

#get top n most prevalent species
top.species.allInfants <- c(species.prevalence[order(desc(species.prevalence$n_tmp)),] %>% 
  head(n=25) %>% 
  pull(Species))

species.list.bothTimepoints.allInfants.topMostPrevs <- intersect(species.list.bothTimepoints.allInfants, top.species.allInfants)

#n iterations
for (i in species.list.bothTimepoints.allInfants.topMostPrevs){

  print(i)
  
  species.target.df <- tmp1 %>% 
    filter(Species==i)
    
  res <- t.test(Abundance ~ Sample_type, data = species.target.df)
  pvalue <- res$p.value

  collect.results.allInfants <- rbind(
         collect.results.allInfants,
         data.frame(Species = i, pvalue = pvalue)
       )
}    

collect.results.allInfants$pvalue.BH <- p.adjust(collect.results.allInfants$pvalue, method = "BH")

mat.all <- collect.results.allInfants %>% 
  select(Species, pvalue.BH) %>% 
  column_to_rownames("Species")

Heatmap(mat.all, name = "1M-6M differential abundance")

#species that have significant pvalues after BH correction
species.signifBHpvalue <- c(mat.all %>%
                              rownames_to_column("Species") %>% 
                              filter(pvalue.BH<0.05) %>% 
                              pull(Species))

pdf(file=paste0("figures/species_stability/boxplots_DAA_1M6M.pdf"), width=7, height=4)
ggplot(tmp1 %>% filter(Species %in% species.signifBHpvalue), aes(x=Timepoint, y=Abundance)) + 
  geom_boxplot(outlier.size = -1)+
  geom_jitter(alpha=0.4, fill="grey", width=0.2)+
  theme_bw()+
  facet_grid(~Species)+
  stat_compare_means(method = "t.test", paired=TRUE)
dev.off()

```


## species stability in infants divided by breastfeeding
```{r}
bf_transition.df <- metadata %>% 
  filter(Sample_type=="Stool_1M" | Sample_type=="Stool_6M") %>% 
  filter(`Mother Infant Pair` %in% couples_bothStool) %>% 
  mutate(Exclusive_bf.num = ifelse(Exclusive_bf=="Yes", "1", "0")) %>% 
  select(`Mother Infant Pair`, Sample_type, Exclusive_bf.num)

bf_transition.df$Exclusive_bf.num <- as.numeric(bf_transition.df$Exclusive_bf.num)

bf_transition.df.categories <- bf_transition.df %>% 
  group_by(`Mother Infant Pair`) %>% 
  summarise(n_exclbf_samples=sum(Exclusive_bf.num)) %>% 
  mutate(bf_transition = ifelse(n_exclbf_samples=="2", "BF-BF", "BF-notBF")) %>% 
  select(-n_exclbf_samples)

merged.species.clean.bfcats <- left_join(merged.species.clean, bf_transition.df.categories, by=c("Mother Infant Pair"))

merged.species.clean.couplesSpecies.bf <- merged.species.clean.bfcats %>% 
  filter((`Mother Infant Pair` %in% couples_bothStool) & (Species %in% top10.species)) %>% 
  filter(!is.na(bf_transition)) %>% 
  mutate(Timepoint = ifelse(Sample_type=="Stool_1M", "1", "6")) %>% 
  group_by(Timepoint, Species, bf_transition) %>% 
  summarise(mean_rel_abs=mean(Abundance)) %>% 
  group_by(Timepoint, bf_transition) %>% 
  mutate(total_rel_abs = sum(mean_rel_abs),
         percentage = mean_rel_abs*100/total_rel_abs)

merged.species.clean.couplesSpecies.bf <- merged.species.clean.couplesSpecies.bf %>% 
  select(Species, Timepoint, percentage, bf_transition)

merged.species.clean.couplesSpecies.bf$Timepoint <- as.numeric(merged.species.clean.couplesSpecies.bf$Timepoint)

bf.stability.plot <- ggplot(merged.species.clean.couplesSpecies.bf, aes(x=Timepoint, y=percentage, fill=Species)) + 
   geom_area(position = 'stack', alpha=0.9, size=.5, colour="black")+
   theme_light() +
   scale_fill_manual(values=rev(custom.stool3)) +
   facet_grid(~bf_transition)

pdf(file=paste0("figures/species_stability/bf_stool_stability.pdf"), width=7, height=4)
bf.stability.plot
invisible(dev.off())
```

## t-test differentially abundant 1M vs 6M
```{r}
tmp2 <- merged.species.clean %>% 
  mutate(Timepoint = ifelse(Sample_type=="Stool_1M", "1", "6")) %>% 
  filter(`Mother Infant Pair` %in% couples_bothStool)

tmp2.bfcat <- left_join(tmp2, bf_transition.df.categories, by=c("Mother Infant Pair"))

#species in both timepoints
species.list.bothTimepoints <- c(tmp2.bfcat %>% 
  group_by(Species, `Mother Infant Pair`) %>% 
  summarise(n=n()) %>% 
  filter(n==2) %>% 
  pull(Species) %>% 
  unique())

#get top n most prevalent species
top.species <- c(species.prevalence[order(desc(species.prevalence$n_tmp)),] %>% 
  head(n=25) %>% 
  pull(Species))

species.list.bothTimepoints.topMostPrevs <- intersect(species.list.bothTimepoints, top.species)

collect.results <- data.frame()
bfcats <- c("BF-BF", "BF-notBF")

for (i in bfcats){
  for (j in species.list.bothTimepoints.topMostPrevs){
  
    print(j)
    
    species.bfcat.target.df <- tmp2.bfcat %>% 
      filter((bf_transition==i) & (Species==j))
      
    res <- t.test(Abundance ~ Timepoint, data = species.bfcat.target.df)
    pvalue <- res$p.value
  
    collect.results <- rbind(
           collect.results,
           data.frame(BF_transition = i, Species = j, pvalue = pvalue)
         )
  }
}    

collect.results$pvalue.BH <- p.adjust(collect.results$pvalue, method = "BH")

mat <- collect.results %>% 
  select(Species, pvalue.BH, BF_transition) 
  
rownames(mat) <- make.names(mat[,1], unique = TRUE) 
Heatmap(mat %>% select(-Species, -BF_transition))
```

#boxplots target taxa
```{r}
#boxplots of target taxa
collect.results$pvalue.BH <- as.numeric(collect.results$pvalue.BH)

#get top taxa for BF-BF results
collect.results.BFBF <- collect.results %>% 
  filter(BF_transition=="BF-BF")
collect.results.BFBF <- collect.results.BFBF[order(collect.results.BFBF$pvalue.BH, decreasing = FALSE),]
collect.results.BFBF.top <- c(collect.results.BFBF %>% 
  slice(1:15) %>% 
  pull(Species))

#get top taxa for BF-notBF results
collect.results.BFnotBF <- collect.results %>% 
  filter(BF_transition=="BF-notBF")
collect.results.BFnotBF <- collect.results.BFnotBF[order(collect.results.BFnotBF$pvalue.BH, decreasing = FALSE),]
collect.results.BFnotBF.top <- c(collect.results.BFnotBF %>% 
  slice(1:5) %>% 
  pull(Species))

target_taxa <- c(collect.results.BFBF.top, collect.results.BFnotBF.top)
target_taxa2 <- c("Bifidobacterium breve", "Bifidobacterium longum", "Escherichia coli", 
                 "Klebsiella pneumoniae", "Ruminococcus gnavus", "Ruminococcus gnavus")

tmp2.bfcat.joinedVar <- tmp2.bfcat %>% 
  filter(!is.na(bf_transition) & Species %in% target_taxa2) %>% 
  mutate(timepoint_transition=paste0(Timepoint, "_", bf_transition)) %>% 
  mutate(timepoint_transition=fct_relevel(timepoint_transition, c("1_BF-BF", "6_BF-BF","1_BF-notBF", "6_BF-notBF")))

p1 <- ggplot(tmp2.bfcat.joinedVar, aes(x=timepoint_transition, y=log10(Abundance), fill=timepoint_transition)) + 
  geom_boxplot(outlier.size = -1, alpha=0.8)+
  #geom_jitter(alpha=0.3, size=0.4, fill="grey", width=0.2)+
  theme_bw()+
  facet_grid(~Species)+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1))+
  scale_fill_manual(values=c("#e66101", "#fdb863", "#5e3c99", "#b2abd2")) 

my_comparisons <- list(c("1_BF-BF", "6_BF-BF"), 
                       c("1_BF-notBF", "6_BF-notBF"),
                       c("1_BF-BF", "1_BF-notBF"),
                       c("6_BF-BF", "6_BF-notBF"))


pdf(file=paste0("figures/species_stability/boxplots_DAA_targetSpecies_pvalues.pdf"), width=10, height=4)
p1 + stat_compare_means(p.adjust.method = "BH", comparisons = my_comparisons, method = "t.test", label='p.adj') 
invisible(dev.off())

```

