---
title: "Beta diversity analysis - PCoA"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)
options(stringsAsFactors = FALSE)
```

#Prepare environment.
```{r}
library(tidyverse)
library(ggplot2)
library("grid")
library("ggrepel") #Avoid overlapping labels
library(stringr)
library(viridis)
library(vegan)
library(cowplot)
library("ggpubr")
```

## Load tables
```{r}
mpa_file <- read_delim("data/species_composition/milk_infant_rel_abundance.txt", delim = '\t')
my_metadata <- read_delim("data/metadata.csv", delim = ',')
bifidos_rel_abs_mean <- read_delim("data/predominance_group/bifidos_relabs_mean.csv", delim = ',')
```

# PCoA plots

## By sample type
```{r}

mpa_file = mpa_file[-1,]
# 
species_t_matrix <- t(mpa_file[,-1])
colnames(species_t_matrix) <- mpa_file$clade_name

#calculate values from scratch
# dist_t_matrix = vegdist(species_t_matrix, method = "bray")
# 
# cmd_t_res = cmdscale(dist_t_matrix, 
#                    k = (nrow(species_t_matrix) - 1),
#                    eig = TRUE)
# 
# str(cmd_t_res)

#saveRDS(cmd_t_res, file="cmd_t_res.RData")

#load pre-computed data
cmd_t_res <- readRDS("cmd_t_res.RData")

#Extracting information of the first two PCs
pcoa_t_df = tibble(PC1 = cmd_t_res$points[,1], 
                 PC2 = cmd_t_res$points[,2])

#plot samples
ggplot(pcoa_t_df, aes(x = PC1, y = PC2)) + 
        geom_point()

#Adding metadata to the plot
ordered_IDs <- tibble::rownames_to_column(as.data.frame(species_t_matrix), "Full_name")
pcoa_t_meta = bind_cols(pcoa_t_df, ordered_IDs[1])
pcoa_t_meta.all = left_join(pcoa_t_meta, my_metadata, by = c("Full_name"="Sample_ID"))

pcoa_t_meta.all.clean <- pcoa_t_meta.all %>% 
  filter(Post_filtering_QC=="Keep")

pcoa_t_meta.all.clean.plot <- ggplot(pcoa_t_meta.all.clean, aes(x = PC1, y = PC2, color = Sample_type)) +
  geom_point(size=2) +
  theme_light() +
  scale_color_manual(values=c("#e66101", "#fdb863", "#5e3c99", "#b2abd2")) +
  theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_sample_type.pdf"), width=6, height=4)
pcoa_t_meta.all.clean.plot
invisible(dev.off())

pcoa_t_meta.all.clean.plot
```


## by mean relative abundance of all bifidobacteria
```{r}

pcoa_t_meta.all.bifidos_mean <- left_join(pcoa_t_meta.all.clean, bifidos_rel_abs_mean, by=c("Correct_sample_name"= "Sample"))

pcoa_t_meta.all.bifidos_mean.clean <- pcoa_t_meta.all.bifidos_mean %>% 
  filter(!is.na(mean_rel_ab)) %>% 
  mutate(mean_rel_ab2 = round(mean_rel_ab,1))
                                        
pcoa_t_meta.all.bifidos_mean.clean.plot <- ggplot(pcoa_t_meta.all.bifidos_mean.clean, aes(x = PC1, y = PC2, color = mean_rel_ab)) +
  geom_point(size=2) +
  theme_light() +
  scale_colour_viridis(option="magma", na.value = "#ffeebf", limits=c(0,40))
  #theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_mean_bifidos_relab_all.pdf"), width=4.5, height=3)
pcoa_t_meta.all.bifidos_mean.clean.plot
invisible(dev.off())

pcoa_t_meta.all.bifidos_mean.clean.plot
```

## By collection time point in infant stool samples only
```{r}
timepoint.stool.plot <- ggplot(pcoa_t_meta.all.clean %>% filter(Bodysite=="Stool"), aes(x = PC1, y = PC2, color = Timepoint)) +
  geom_point(size=2) +
  theme_light() +
  scale_color_manual(values=c("navy", "darkorange2"))+
  theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_timepoint.pdf"), width=width_size, height=height_size)
timepoint.stool.plot
invisible(dev.off())

timepoint.stool.plot
```

## By pre-partum antibiotic intake
```{r}
prenatal.abx.plot <- ggplot(pcoa_t_meta.all.clean, aes(x = PC1, y = PC2, color = `Prenatal_abx`)) +
  geom_point(size=2) +
  theme_light() +
  scale_color_manual(values=c("darkorange2", "black"))+
  theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_prenatal_abx.pdf"), width=width_size, height=height_size)
prenatal.abx.plot
invisible(dev.off())

prenatal.abx.plot
```

## By overall antibiotics intake 
```{r}

abx.any.plot <- ggplot(pcoa_t_meta.all.clean %>% filter(Bodysite=="Stool"), aes(x = PC1, y = PC2, color = `Abx_exposure_any`)) +
  geom_point(size=2) +
  theme_light() +
  scale_color_manual(values=c("darkorange2", "black"))+
  theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_abx_any.pdf"), width=width_size, height=height_size)
abx.any.plot
invisible(dev.off())

abx.any.plot
```

## By delivery mode
```{r}
delivery.mode.clean <- pcoa_t_meta.all.clean %>% 
  filter(Bodysite=="Stool") %>% 
  mutate(Delivery_mode=ifelse(Delivery_mode_cat1=="AVD"|Delivery_mode_cat1=="SVD", "Vaginal",
                              ifelse(Delivery_mode_cat1=="CS", "Ceserean", 
                                     ifelse(Delivery_mode_cat1=="0", NA, Delivery_mode_cat1))))

delivery.mode.clean.plot <- ggplot(delivery.mode.clean, aes(x = PC1, y = PC2, color = `Delivery_mode`)) +
  geom_point(size=2) +
  theme_light() +
  scale_color_manual(values=c("darkorange2", "black"))+
  theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_delivery_mode.pdf"), width=5, height=height_size)
delivery.mode.clean.plot
invisible(dev.off())

delivery.mode.clean.plot
```

## By place of birth
```{r}
hospital.plot <- ggplot(pcoa_t_meta.all.clean %>% filter(!is.na(Place_birth)), aes(x = PC1, y = PC2, color = `Place_birth`)) +
  geom_point(size=2) +
  theme_light() +
  scale_color_manual(values=c("skyblue","darkgreen","#b2df8a","purple","pink",
                                       "violetred2","red","darkorange","gold", "navy"))+
  theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_hospital.pdf"), width=6, height=height_size)
hospital.plot
invisible(dev.off())

hospital.plot
```

## By feeding mode at 6 months of age 
```{r}
excl.bf.6m.plot <- ggplot(pcoa_t_meta.all.clean %>% filter(Bodysite=="Stool"), aes(x = PC1, y = PC2, color = `Exclusive_bf_6m`)) +
  geom_point(size=2) +
  theme_light() +
  scale_color_manual(values=c("darkorange2", "black"))+
  theme(legend.position="bottom")

pdf(file=paste0("figures/beta_diversity/PCoA_exc_bf_6m.pdf"), width=width_size, height=height_size)
excl.bf.6m.plot
invisible(dev.off())

excl.bf.6m.plot
```

## create panel of all PCoA plots
```{r}
panel <- plot_grid(cum.rel.ab.bifiods.plot, delivery.mode.clean.plot,excl.bf.6m.plot, timepoint.stool.plot, abx.any.plot, hospital.plot, labels = "auto", ncol=2)

pdf(file=paste0("figures/beta_diversity/cum_bifidos_relab.pdf"), width=8, height=10)
panel
dev.off()

panel
```

## Biplots: Add species information to the PCoA

Biplot indicating which species are driving the variation in the taxonomic composition by using the Weighted Averages Scores for Species: wascores()
see https://github.com/biobakery/biobakery/wiki/Metagenomic-Visualizations#pcoas-and-biplots for reference 

```{r}

mpa_mat = mpa_file |> 
  column_to_rownames("clade_name") |> 
  as.matrix() |>
  t()

wa_data = wascores(cmd_t_res$points[,1:2], mpa_mat) |>
    as_tibble(rownames = 'species')

cmd_t_res <- readRDS("cmd_t_res.RData")

#Extracting information of the first two PCs
pcoa_t_df = tibble(PC1 = cmd_t_res$points[,1], 
                 PC2 = cmd_t_res$points[,2])

#wa_data

#too many species to plot, selected those of interest
#species of interest were the ten most prevalent species found in infant stool samples at 1 month of age 
species_list = c("k__Bacteria|p__Actinobacteria|c__Actinobacteria|o__Bifidobacteriales|f__Bifidobacteriaceae|g__Bifidobacterium|s__Bifidobacterium_longum", 
                 "k__Bacteria|p__Actinobacteria|c__Actinobacteria|o__Bifidobacteriales|f__Bifidobacteriaceae|g__Bifidobacterium|s__Bifidobacterium_breve", 
                 "k__Bacteria|p__Actinobacteria|c__Actinobacteria|o__Bifidobacteriales|f__Bifidobacteriaceae|g__Bifidobacterium|s__Bifidobacterium_bifidum", 
                 "k__Bacteria|p__Proteobacteria|c__Gammaproteobacteria|o__Enterobacterales|f__Enterobacteriaceae|g__Escherichia|s__Escherichia_coli", 
                 "k__Bacteria|p__Bacteroidetes|c__Bacteroidia|o__Bacteroidales|f__Bacteroidales_unclassified|g__Phocaeicola|s__Phocaeicola_vulgatus", 
                 "k__Bacteria|p__Bacteroidetes|c__Bacteroidia|o__Bacteroidales|f__Bacteroidaceae|g__Bacteroides|s__Bacteroides_fragilis", 
                 "k__Bacteria|p__Proteobacteria|c__Gammaproteobacteria|o__Enterobacterales|f__Enterobacteriaceae|g__Klebsiella|s__Klebsiella_pneumoniae",
                 "k__Bacteria|p__Proteobacteria|c__Gammaproteobacteria|o__Enterobacterales|f__Enterobacteriaceae|g__Klebsiella|s__Klebsiella_michiganensis", 
                 "k__Bacteria|p__Firmicutes|c__Negativicutes|o__Veillonellales|f__Veillonellaceae|g__Veillonella|s__Veillonella_atypica", 
                 "k__Bacteria|p__Firmicutes|c__Negativicutes|o__Veillonellales|f__Veillonellaceae|g__Veillonella|s__Veillonella_dispar",
                 "k__Bacteria|p__Firmicutes|c__Negativicutes|o__Veillonellales|f__Veillonellaceae|g__Veillonella|s__Veillonella_parvula")
  
#shorten species name
wa_species_subset = wa_data %>%  
  filter(species %in% species_list) %>% 
  mutate(species_clean = case_when(
    grepl("breve", species) ~ "B. breve",
    grepl("longum", species) ~ "B. longum",
    grepl("bifidum", species) ~ "B. bifidium",
    grepl("coli", species) ~ "E. coli",
    grepl("fragilis", species) ~ "B. fragilis",
    grepl("pneumoniae", species) ~ "K. pneumoniae",
    grepl("michiganensis", species) ~ "K. michiganensis",
    grepl("atypica", species) ~ "V. atypica",
    grepl("dispar", species) ~ "V. dispar",
    grepl("parvula", species) ~ "V. parvula",
    grepl("vulgatus", species, ignore.case = TRUE) ~"P. vulgatus"))
  
pcoa.labels.plot <- ggplot(pcoa_t_meta.all.clean, aes(x = PC1, y = PC2, color = Sample_type)) +
  geom_point(size=1.8) +
  theme_light() +
  scale_color_manual(values=c("#e66101", "#fdb863", "#5e3c99", "#b2abd2")) +
  geom_text(data = wa_species_subset,
              aes(x = V1, y = V2, 
                  label = species_clean), 
                  color = "black", 
                  size = 5)

pdf(file=paste0("figures/beta_diversity/PCoA_biplot_species.pdf"), width=5.1, height=height_size)
pcoa.labels.plot
invisible(dev.off())

pcoa.labels.plot
```
